---
title: "DGE"
format: html
editor: visual
---

```{r}
# ============================================================
# nf-core/rnaseq gene-level differential expression analysis
# ============================================================

# Install once if needed:
# if (!requireNamespace("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
#
# BiocManager::install(c("DESeq2", "pheatmap", "rtracklayer"))
# install.packages(c("ggplot2", "ggrepel", "matrixStats"))
#
# Optional, but recommended for LFC shrinkage:
# BiocManager::install("ashr")

library(DESeq2)
library(ggplot2)
library(ggrepel)
library(pheatmap)
library(matrixStats)
```


```{r}
# ----------------------------
# 1. Import count matrix
# ----------------------------

count_file <- "salmon.merged.gene_counts.tsv"

dat <- read.delim(
  count_file,
  header = TRUE,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

stopifnot(all(c("gene_id", "gene_name") %in% names(dat)))
stopifnot(!anyDuplicated(dat$gene_id))

gene_annotation <- dat[, c("gene_id", "gene_name")]

# Add product descriptions from the EGAPx GTF. Product annotations are
# transcript/isoform-level and are repeated on exon/CDS records, so prefer
# transcript records and collapse distinct isoform products once per gene.
gtf_file <- "complete.genomic.gtf"

if (!requireNamespace("rtracklayer", quietly = TRUE)) {
  stop(
    "Package 'rtracklayer' is required. Install it with: ",
    "BiocManager::install('rtracklayer')"
  )
}

gtf <- rtracklayer::import(gtf_file)

if (!all(c("gene_id", "type", "product") %in% names(mcols(gtf)))) {
  stop("The GTF must contain gene_id, type, and product attributes.")
}

gtf_products <- data.frame(
  gene_id = as.character(mcols(gtf)$gene_id),
  feature_type = as.character(mcols(gtf)$type),
  product = as.character(mcols(gtf)$product),
  stringsAsFactors = FALSE
)

gtf_products <- gtf_products[
  !is.na(gtf_products$gene_id) &
    gtf_products$gene_id != "" &
    !is.na(gtf_products$product) &
    gtf_products$product != "",
]

product_annotation <- do.call(
  rbind,
  lapply(split(gtf_products, gtf_products$gene_id), function(x) {
    transcript_products <- unique(
      x$product[x$feature_type %in% c("transcript", "mRNA")]
    )

    products <- if (length(transcript_products) > 0) {
      transcript_products
    } else {
      unique(x$product)
    }

    data.frame(
      gene_id = x$gene_id[1],
      product_description = paste(products, collapse = " | "),
      stringsAsFactors = FALSE
    )
  })
)

rownames(product_annotation) <- NULL

gene_annotation$product_description <- product_annotation$product_description[
  match(gene_annotation$gene_id, product_annotation$gene_id)
]

message(
  sum(!is.na(gene_annotation$product_description)),
  " of ",
  nrow(gene_annotation),
  " count-table genes matched to a GTF product description."
)

rm(gtf, gtf_products, product_annotation)

count_data <- as.matrix(
  dat[, setdiff(names(dat), c("gene_id", "gene_name"))]
)

rownames(count_data) <- dat$gene_id
storage.mode(count_data) <- "numeric"

# DESeq2 requires nonnegative count-like input.
stopifnot(!anyNA(count_data))
stopifnot(all(count_data >= 0))

# The supplied merged file appears integer-valued. This also handles
# very small floating-point deviations if estimated counts are present.
if (!all(count_data == round(count_data))) {
  message("Non-integer estimated counts found; rounding for direct DESeq2 import.")
  count_data <- round(count_data)
}

# Remove annotation table from the working environment if desired.
rm(dat)
```


```{r}
# ----------------------------
# 2. Construct sample metadata
# ----------------------------

sample_id <- colnames(count_data)
sample_core <- sub("_S[0-9]+_L[0-9]+$", "", sample_id)

coldata <- data.frame(
  sample_id = sample_id,
  sample_core = sample_core,
  stringsAsFactors = FALSE
)

rownames(coldata) <- coldata$sample_id

coldata$day <- ifelse(
  grepl("^[CF]45-", coldata$sample_core), 45L,
  ifelse(grepl("^[CF]103-", coldata$sample_core), 103L, NA_integer_)
)

coldata$stage <- factor(
  ifelse(coldata$day == 45, "Young", "Old"),
  levels = c("Young", "Old")
)

coldata$treatment <- factor(
  ifelse(grepl("^F", coldata$sample_core), "Flutamide", "Control"),
  levels = c("Control", "Flutamide")
)

coldata$tissue <- factor(
  ifelse(grepl("B$", coldata$sample_core), "Body",
         ifelse(grepl("H$", coldata$sample_core), "Head", NA)),
  levels = c("Body", "Head")
)

# Tadpole identifier, retained for plotting and QC.
# Do not include it in the DGE model: tadpoles are nested within
# stage/treatment, so subject and group would be confounded.
coldata$tadpole <- sub("[BH]$", "", coldata$sample_core)

stopifnot(!anyNA(coldata[, c("day", "stage", "treatment", "tissue")]))
stopifnot(identical(rownames(coldata), colnames(count_data)))

# Combined experimental group. Using one factor makes the requested
# contrasts explicit and avoids unnecessary interaction complexity.
coldata$group <- factor(
  paste(coldata$stage, coldata$treatment, coldata$tissue, sep = "_"),
  levels = c(
    "Young_Control_Body",
    "Young_Control_Head",
    "Old_Control_Body",
    "Old_Control_Head",
    "Old_Flutamide_Body",
    "Old_Flutamide_Head"
  )
)

# Check replication and metadata assignment.
print(coldata[, c(
  "sample_core", "tadpole", "day",
  "stage", "treatment", "tissue", "group"
)])

print(table(coldata$group))

# Expected:
# Young_Control_Body     5
# Young_Control_Head     4
# Old_Control_Body       4
# Old_Control_Head       3
# Old_Flutamide_Body     5
# Old_Flutamide_Head     4
```


```{r}
# ----------------------------
# 3. Build and fit DESeq2 model
# ----------------------------

dds <- DESeqDataSetFromMatrix(
  countData = count_data,
  colData = coldata,
  design = ~ group
)

# Attach gene names without using them as row identifiers.
# Gene names may not be unique; gene_id is the safer row identifier.
mcols(dds)$gene_name <- gene_annotation$gene_name[
  match(rownames(dds), gene_annotation$gene_id)
]

mcols(dds)$product_description <- gene_annotation$product_description[
  match(rownames(dds), gene_annotation$gene_id)
]

# Bulk RNA-seq prefilter recommended in the DESeq2 manual:
# count >= 10 in at least the smallest group size (3 samples).
keep <- rowSums(counts(dds) >= 10) >= 3
message(sum(keep), " of ", length(keep), " genes retained.")
dds <- dds[keep, ]

dds <- DESeq(dds)

resultsNames(dds)
plotDispEsts(dds)

# Create output directories.
dir.create("DESeq2_results", showWarnings = FALSE)
dir.create("DESeq2_results/QC", showWarnings = FALSE)
dir.create("DESeq2_results/DGE", showWarnings = FALSE)
dir.create("DESeq2_results/volcano", showWarnings = FALSE)

write.csv(
  as.data.frame(colData(dds)),
  "DESeq2_results/sample_metadata.csv",
  row.names = FALSE
)
```


```{r}
# ----------------------------
# 4. VST for PCA and clustering
# ----------------------------

# blind=FALSE is appropriate for biological visualization when large,
# expected tissue/stage effects exist. It uses the fitted dispersion trend;
# it does not regress the group effects out of the transformed matrix.
vsd <- vst(dds, blind = FALSE)
vst_matrix <- assay(vsd)

# Select the 500 genes with greatest variance across samples.
gene_variance <- rowVars(vst_matrix)
names(gene_variance) <- rownames(vst_matrix)

n_top <- min(500, nrow(vst_matrix))
top500 <- names(sort(gene_variance, decreasing = TRUE))[seq_len(n_top)]

write.csv(
  data.frame(
    gene_id = top500,
    gene_name = mcols(dds)[top500, "gene_name"],
    VST_variance = gene_variance[top500]
  ),
  "DESeq2_results/QC/top500_variable_genes.csv",
  row.names = FALSE
)
```


```{r}
# ----------------------------
# 5. PCA: top 500 variable genes
# ----------------------------

pca <- prcomp(t(vst_matrix[top500, , drop = FALSE]), scale. = FALSE)

percent_var <- 100 * pca$sdev^2 / sum(pca$sdev^2)

pca_data <- cbind(
  as.data.frame(pca$x[, 1:2, drop = FALSE]),
  as.data.frame(colData(vsd))
)

pca_plot <- ggplot(
  pca_data,
  aes(
    x = PC1,
    y = PC2,
    color = group,
    shape = tissue,
    label = sample_core
  )
) +
  geom_point(size = 3.5) +
  geom_text_repel(
    size = 3,
    max.overlaps = Inf,
    show.legend = FALSE
  ) +
  labs(
    title = "PCA of all samples",
    subtitle = paste0(
      "VST-transformed counts; ",
      n_top,
      " most variable genes"
    ),
    x = sprintf("PC1 (%.1f%%)", percent_var[1]),
    y = sprintf("PC2 (%.1f%%)", percent_var[2]),
    color = "Experimental group",
    shape = "Tissue"
  ) +
  coord_fixed() +
  theme_classic(base_size = 12) +
  theme(legend.position = "right")

ggsave(
  "DESeq2_results/QC/PCA_top500_variable_genes.pdf",
  pca_plot,
  width = 10,
  height = 7
)

ggsave(
  "DESeq2_results/QC/PCA_top500_variable_genes.png",
  pca_plot,
  width = 10,
  height = 7,
  dpi = 300
)

write.csv(
  pca_data,
  "DESeq2_results/QC/PCA_sample_coordinates.csv",
  row.names = FALSE
)
```


```{r}
# ----------------------------
# 6. Sample-clustering dendrogram
# ----------------------------

# Euclidean distance across the same top 500 VST genes.
sample_dist <- dist(t(vst_matrix[top500, , drop = FALSE]))
sample_hc <- hclust(sample_dist, method = "complete")

sample_labels <- paste(
  coldata[sample_hc$labels, "sample_core"],
  coldata[sample_hc$labels, "group"],
  sep = " | "
)

pdf(
  "DESeq2_results/QC/sample_dendrogram_top500.pdf",
  width = 13,
  height = 7
)

plot(
  sample_hc,
  labels = sample_labels,
  main = "Hierarchical clustering of samples",
  sub = paste(n_top, "most variable VST-transformed genes"),
  xlab = "",
  ylab = "Euclidean distance",
  cex = 0.7,
  hang = -1
)

dev.off()

# Sample-to-sample distance heatmap.
sample_dist_matrix <- as.matrix(sample_dist)

annotation_col <- as.data.frame(
  colData(vsd)[, c("stage", "treatment", "tissue", "group")]
)

pheatmap(
  sample_dist_matrix,
  annotation_col = annotation_col,
  annotation_row = annotation_col,
  clustering_distance_rows = sample_dist,
  clustering_distance_cols = sample_dist,
  filename = "DESeq2_results/QC/sample_distance_heatmap.pdf",
  width = 11,
  height = 10,
  main = "Sample distances: top 500 variable genes"
)
```


```{r}
# ----------------------------
# 7. Gene-expression heatmap
# ----------------------------

# For a readable gene-level heatmap, use the top 50 rather than all genes.
top50 <- top500[seq_len(min(50, length(top500)))]

heatmap_matrix <- vst_matrix[top50, , drop = FALSE]

# Row-wise z-scores show relative expression patterns per gene.
heatmap_z <- t(scale(t(heatmap_matrix)))
heatmap_z[!is.finite(heatmap_z)] <- 0

gene_labels <- mcols(dds)[top50, "gene_name"]
gene_labels <- ifelse(
  is.na(gene_labels) | gene_labels == "",
  top50,
  paste0(gene_labels, " [", top50, "]")
)

rownames(heatmap_z) <- make.unique(gene_labels)

pheatmap(
  heatmap_z,
  annotation_col = annotation_col,
  show_colnames = TRUE,
  show_rownames = TRUE,
  fontsize_row = 6,
  fontsize_col = 7,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  clustering_method = "complete",
  color = colorRampPalette(c("navy", "white", "firebrick3"))(101),
  breaks = seq(-3, 3, length.out = 102),
  filename = "DESeq2_results/QC/top50_variable_gene_heatmap.pdf",
  width = 12,
  height = 12,
  main = "Top 50 variable genes (row z-scores)"
)
```


```{r}
# ----------------------------
# 8. DGE and volcano function
# ----------------------------

run_comparison <- function(
  dds,
  numerator,
  denominator,
  comparison_name,
  padj_cutoff = 0.05,
  lfc_cutoff = 1
) {
  contrast_vector <- c("group", numerator, denominator)

  # Statistical results use the DESeq2 Wald test.
  res <- results(
    dds,
    contrast = contrast_vector,
    alpha = padj_cutoff
  )

  # Shrink LFCs for ranking and visualization.
  # ashr preserves large effects and accepts arbitrary contrasts.
  # If unavailable, DESeq2's built-in normal shrinker is used.
  if (requireNamespace("ashr", quietly = TRUE)) {
    res_shrunk <- lfcShrink(
      dds,
      contrast = contrast_vector,
      res = res,
      type = "ashr"
    )
    shrink_method <- "ashr"
  } else {
    warning(
      "Package 'ashr' is unavailable; using DESeq2 type='normal'. ",
      "Install ashr for the preferred wide-tailed shrinkage estimator."
    )
    res_shrunk <- lfcShrink(
      dds,
      contrast = contrast_vector,
      res = res,
      type = "normal"
    )
    shrink_method <- "normal"
  }

  out <- as.data.frame(res)
  out$gene_id <- rownames(out)
  out$gene_name <- mcols(dds)[out$gene_id, "gene_name"]
  out$product_description <- mcols(dds)[
    out$gene_id,
    "product_description"
  ]
  out$shrunken_log2FoldChange <- res_shrunk$log2FoldChange
  out$shrunken_lfcSE <- res_shrunk$lfcSE

  # Put identifiers first.
  out <- out[, c(
    "gene_id",
    "gene_name",
    "product_description",
    "baseMean",
    "log2FoldChange",
    "lfcSE",
    "shrunken_log2FoldChange",
    "shrunken_lfcSE",
    "stat",
    "pvalue",
    "padj"
  )]

  out <- out[order(out$padj, out$pvalue, na.last = TRUE), ]

  out$significance <- "Not significant"
  out$significance[
    !is.na(out$padj) &
      out$padj < padj_cutoff &
      out$shrunken_log2FoldChange >= lfc_cutoff
  ] <- "Up"

  out$significance[
    !is.na(out$padj) &
      out$padj < padj_cutoff &
      out$shrunken_log2FoldChange <= -lfc_cutoff
  ] <- "Down"

  write.csv(
    out,
    file.path(
      "DESeq2_results/DGE",
      paste0(comparison_name, "_DESeq2_results.csv")
    ),
    row.names = FALSE
  )

  # For plotting only, NA adjusted p-values are treated as nonsignificant.
  # DESeq2 may produce NA padj for low-information genes or outliers.
  plot_data <- out
  plot_data$plot_padj <- ifelse(
    is.na(plot_data$padj),
    1,
    pmax(plot_data$padj, .Machine$double.xmin)
  )

  # Label the 10 most significant genes meeting both thresholds.
  label_candidates <- subset(
    plot_data,
    significance != "Not significant" &
      !is.na(gene_name) &
      gene_name != ""
  )

  label_candidates <- head(
    label_candidates[order(label_candidates$padj), ],
    10
  )

  volcano <- ggplot(
    plot_data,
    aes(
      x = shrunken_log2FoldChange,
      y = -log10(plot_padj),
      color = significance
    )
  ) +
    geom_point(alpha = 0.65, size = 1.5) +
    geom_vline(
      xintercept = c(-lfc_cutoff, lfc_cutoff),
      linetype = "dashed",
      color = "grey40"
    ) +
    geom_hline(
      yintercept = -log10(padj_cutoff),
      linetype = "dashed",
      color = "grey40"
    ) +
    geom_text_repel(
      data = label_candidates,
      aes(label = gene_name),
      size = 3,
      box.padding = 0.4,
      max.overlaps = Inf,
      show.legend = FALSE
    ) +
    scale_color_manual(
      values = c(
        "Down" = "#2C7BB6",
        "Not significant" = "grey75",
        "Up" = "#D7191C"
      ),
      breaks = c("Up", "Down", "Not significant")
    ) +
    labs(
      title = comparison_name,
      subtitle = paste0(
        numerator,
        " vs ",
        denominator,
        "; LFC shrinkage: ",
        shrink_method
      ),
      x = "Shrunken log2 fold change",
      y = expression(-log[10]("adjusted p-value")),
      color = NULL,
      caption = paste0(
        "Significant: adjusted p < ",
        padj_cutoff,
        " and |shrunken log2FC| \u2265 ",
        lfc_cutoff
      )
    ) +
    theme_classic(base_size = 12) +
    theme(legend.position = "top")

  ggsave(
    file.path(
      "DESeq2_results/volcano",
      paste0(comparison_name, "_volcano.pdf")
    ),
    volcano,
    width = 8,
    height = 7
  )

  ggsave(
    file.path(
      "DESeq2_results/volcano",
      paste0(comparison_name, "_volcano.png")
    ),
    volcano,
    width = 8,
    height = 7,
    dpi = 300
  )

  message(
    comparison_name,
    ": ",
    sum(out$significance == "Up", na.rm = TRUE),
    " up and ",
    sum(out$significance == "Down", na.rm = TRUE),
    " down"
  )

  return(out)
}
```


```{r}
# ----------------------------
# 9. Requested comparisons
# ----------------------------

# Positive LFC = higher expression in OLD control body.
leg_emergence <- run_comparison(
  dds = dds,
  numerator = "Old_Control_Body",
  denominator = "Young_Control_Body",
  comparison_name = "leg_emergence_OldControlBody_vs_YoungControlBody"
)

# Positive LFC = higher expression in OLD control head.
nervous_system_development <- run_comparison(
  dds = dds,
  numerator = "Old_Control_Head",
  denominator = "Young_Control_Head",
  comparison_name = "nervous_system_OldControlHead_vs_YoungControlHead"
)

# Positive LFC = higher expression after FLUTAMIDE treatment.
ar_blockade_body <- run_comparison(
  dds = dds,
  numerator = "Old_Flutamide_Body",
  denominator = "Old_Control_Body",
  comparison_name = "AR_blockade_OldBodyFlutamide_vs_OldBodyControl"
)

# Positive LFC = higher expression after FLUTAMIDE treatment in head tissue.
ar_blockade_head <- run_comparison(
  dds = dds,
  numerator = "Old_Flutamide_Head",
  denominator = "Old_Control_Head",
  comparison_name = "AR_blockade_OldHeadFlutamide_vs_OldHeadControl"
)

```


```{r}
# ----------------------------
# 10. Older-only AR-blockade tissue interaction
# ----------------------------

# Formally test whether the flutamide effect differs between head and body:
#
# (Flutamide head - Control head) -
# (Flutamide body - Control body)
#
# Head and body were collected from the same tadpoles when both libraries
# were available. Including tadpole as a fixed effect accounts for this
# pairing. Tadpole absorbs the between-animal treatment effect, while the
# flutamide_head coefficient estimates the treatment-by-tissue interaction.

dds_old <- dds[, dds$stage == "Old"]

dds_old$tadpole <- factor(dds_old$tadpole)
dds_old$tissue <- factor(
  dds_old$tissue,
  levels = c("Body", "Head")
)
dds_old$treatment <- factor(
  dds_old$treatment,
  levels = c("Control", "Flutamide")
)

# Numeric indicator for the flutamide-by-head interaction.
dds_old$flutamide_head <- as.integer(
  dds_old$treatment == "Flutamide" &
    dds_old$tissue == "Head"
)

design(dds_old) <- ~ tadpole + tissue + flutamide_head

# Stop with an informative error if the paired model is not estimable.
old_model_matrix <- model.matrix(
  design(dds_old),
  data = as.data.frame(colData(dds_old))
)

if (qr(old_model_matrix)$rank != ncol(old_model_matrix)) {
  stop("The older-only paired interaction model is not full rank.")
}

message(
  "Older-only interaction model: ",
  ncol(dds_old),
  " libraries from ",
  nlevels(dds_old$tadpole),
  " tadpoles; residual degrees of freedom = ",
  nrow(old_model_matrix) - qr(old_model_matrix)$rank,
  "."
)

# Re-estimate dispersions and coefficients for the older-only paired model.
dds_old <- DESeq(dds_old)

interaction_name <- grep(
  "flutamide_head",
  resultsNames(dds_old),
  value = TRUE
)

stopifnot(length(interaction_name) == 1)

res_interaction <- results(
  dds_old,
  name = interaction_name,
  alpha = 0.05
)

# Use the same ashr shrinkage estimator as the other DGE comparisons so
# effect-size estimates and volcano plots are methodologically consistent.
if (requireNamespace("ashr", quietly = TRUE)) {
  res_interaction_shrunk <- lfcShrink(
    dds_old,
    coef = interaction_name,
    res = res_interaction,
    type = "ashr"
  )
  interaction_shrink_method <- "ashr"
} else {
  warning(
    "Package 'ashr' is unavailable; the interaction CSV and plot will ",
    "use unshrunken log2 fold changes. Install it with ",
    "BiocManager::install('ashr')."
  )
  res_interaction_shrunk <- res_interaction
  interaction_shrink_method <- "none"
}

# Retain the two component contrasts to make the interaction interpretable.
res_old_head <- results(
  dds,
  contrast = c(
    "group",
    "Old_Flutamide_Head",
    "Old_Control_Head"
  ),
  alpha = 0.05
)

res_old_body <- results(
  dds,
  contrast = c(
    "group",
    "Old_Flutamide_Body",
    "Old_Control_Body"
  ),
  alpha = 0.05
)

interaction_out <- as.data.frame(res_interaction)
interaction_out$gene_id <- rownames(interaction_out)
interaction_out$gene_name <- mcols(dds_old)[
  interaction_out$gene_id,
  "gene_name"
]
interaction_out$product_description <- mcols(dds_old)[
  interaction_out$gene_id,
  "product_description"
]
interaction_out$head_flutamide_log2FC <- res_old_head$log2FoldChange[
  match(interaction_out$gene_id, rownames(res_old_head))
]
interaction_out$head_flutamide_padj <- res_old_head$padj[
  match(interaction_out$gene_id, rownames(res_old_head))
]
interaction_out$body_flutamide_log2FC <- res_old_body$log2FoldChange[
  match(interaction_out$gene_id, rownames(res_old_body))
]
interaction_out$body_flutamide_padj <- res_old_body$padj[
  match(interaction_out$gene_id, rownames(res_old_body))
]
interaction_out$shrunken_interaction_log2FC <-
  res_interaction_shrunk$log2FoldChange
interaction_out$shrunken_interaction_lfcSE <-
  res_interaction_shrunk$lfcSE

interaction_out <- interaction_out[, c(
  "gene_id",
  "gene_name",
  "product_description",
  "baseMean",
  "head_flutamide_log2FC",
  "head_flutamide_padj",
  "body_flutamide_log2FC",
  "body_flutamide_padj",
  "log2FoldChange",
  "lfcSE",
  "shrunken_interaction_log2FC",
  "shrunken_interaction_lfcSE",
  "stat",
  "pvalue",
  "padj"
)]

interaction_out <- interaction_out[
  order(interaction_out$padj, interaction_out$pvalue, na.last = TRUE),
]

interaction_out$interaction_result <- "Not significant"
interaction_out$interaction_result[
  !is.na(interaction_out$padj) &
    interaction_out$padj < 0.05 &
    interaction_out$shrunken_interaction_log2FC > 0
] <- "Positive head-minus-body interaction"
interaction_out$interaction_result[
  !is.na(interaction_out$padj) &
    interaction_out$padj < 0.05 &
    interaction_out$shrunken_interaction_log2FC < 0
] <- "Negative head-minus-body interaction"

write.csv(
  interaction_out,
  paste0(
    "DESeq2_results/DGE/",
    "AR_blockade_head_vs_body_interaction_old_only.csv"
  ),
  row.names = FALSE
)

summary(res_interaction)

# Interaction volcano plot.
interaction_plot_data <- interaction_out
interaction_plot_data$plot_padj <- ifelse(
  is.na(interaction_plot_data$padj),
  1,
  pmax(interaction_plot_data$padj, .Machine$double.xmin)
)

interaction_labels <- subset(
  interaction_plot_data,
  interaction_result != "Not significant" &
    !is.na(gene_name) &
    gene_name != ""
)
interaction_labels <- head(
  interaction_labels[order(interaction_labels$padj), ],
  10
)

interaction_volcano <- ggplot(
  interaction_plot_data,
  aes(
    x = shrunken_interaction_log2FC,
    y = -log10(plot_padj),
    color = interaction_result
  )
) +
  geom_point(alpha = 0.65, size = 1.5) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed",
    color = "grey40"
  ) +
  geom_text_repel(
    data = interaction_labels,
    aes(label = gene_name),
    size = 3,
    max.overlaps = Inf,
    show.legend = FALSE
  ) +
  scale_color_manual(
    values = c(
      "Negative head-minus-body interaction" = "#2C7BB6",
      "Not significant" = "grey75",
      "Positive head-minus-body interaction" = "#D7191C"
    )
  ) +
  labs(
    title = "Older tadpoles: AR-blockade tissue interaction",
    subtitle = paste0(
      "(Flutamide head - control head) - ",
      "(flutamide body - control body); shrinkage: ",
      interaction_shrink_method
    ),
    x = "Head-minus-body interaction log2 fold change",
    y = expression(-log[10]("interaction adjusted p-value")),
    color = NULL
  ) +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

ggsave(
  "DESeq2_results/volcano/AR_blockade_head_vs_body_interaction_old_only.pdf",
  interaction_volcano,
  width = 9,
  height = 7
)

ggsave(
  "DESeq2_results/volcano/AR_blockade_head_vs_body_interaction_old_only.png",
  interaction_volcano,
  width = 9,
  height = 7,
  dpi = 300
)

```


```{r}
# ----------------------------
# 11. Unpaired interaction sensitivity analysis
# ----------------------------

# This model uses all older head and body libraries but treats them as
# independent samples. It is included as a sensitivity analysis; the paired
# tadpole-adjusted interaction above remains the primary analysis.

dds_old_unpaired <- dds[, dds$stage == "Old"]
dds_old_unpaired$treatment <- factor(
  dds_old_unpaired$treatment,
  levels = c("Control", "Flutamide")
)
dds_old_unpaired$tissue <- factor(
  dds_old_unpaired$tissue,
  levels = c("Body", "Head")
)

design(dds_old_unpaired) <- ~ treatment * tissue

unpaired_model_matrix <- model.matrix(
  design(dds_old_unpaired),
  data = as.data.frame(colData(dds_old_unpaired))
)

if (qr(unpaired_model_matrix)$rank != ncol(unpaired_model_matrix)) {
  stop("The older-only unpaired interaction model is not full rank.")
}

dds_old_unpaired <- DESeq(dds_old_unpaired)

unpaired_interaction_name <- grep(
  "treatment.*tissue|tissue.*treatment",
  resultsNames(dds_old_unpaired),
  value = TRUE
)

stopifnot(length(unpaired_interaction_name) == 1)

res_interaction_unpaired <- results(
  dds_old_unpaired,
  name = unpaired_interaction_name,
  alpha = 0.05
)

if (requireNamespace("ashr", quietly = TRUE)) {
  res_interaction_unpaired_shrunk <- lfcShrink(
    dds_old_unpaired,
    coef = unpaired_interaction_name,
    res = res_interaction_unpaired,
    type = "ashr"
  )
} else {
  warning(
    "Package 'ashr' is unavailable; the unpaired sensitivity analysis ",
    "will use unshrunken log2 fold changes."
  )
  res_interaction_unpaired_shrunk <- res_interaction_unpaired
}

# Construct a gene-level comparison in the same orientation for both models:
# (Flutamide head - Control head) -
# (Flutamide body - Control body).
sensitivity_comparison <- data.frame(
  gene_id = rownames(res_interaction),
  gene_name = mcols(dds_old)[rownames(res_interaction), "gene_name"],
  product_description = mcols(dds_old)[
    rownames(res_interaction),
    "product_description"
  ],
  paired_interaction_log2FC = res_interaction_shrunk$log2FoldChange,
  paired_interaction_padj = res_interaction$padj,
  unpaired_interaction_log2FC =
    res_interaction_unpaired_shrunk$log2FoldChange[
      match(
        rownames(res_interaction),
        rownames(res_interaction_unpaired_shrunk)
      )
    ],
  unpaired_interaction_padj = res_interaction_unpaired$padj[
    match(
      rownames(res_interaction),
      rownames(res_interaction_unpaired)
    )
  ],
  stringsAsFactors = FALSE
)

sensitivity_comparison$direction_agrees <- with(
  sensitivity_comparison,
  ifelse(
    is.na(paired_interaction_log2FC) |
      is.na(unpaired_interaction_log2FC),
    NA,
    sign(paired_interaction_log2FC) ==
      sign(unpaired_interaction_log2FC)
  )
)

sensitivity_comparison$paired_significant <- with(
  sensitivity_comparison,
  !is.na(paired_interaction_padj) & paired_interaction_padj < 0.05
)

sensitivity_comparison$unpaired_significant <- with(
  sensitivity_comparison,
  !is.na(unpaired_interaction_padj) & unpaired_interaction_padj < 0.05
)

sensitivity_comparison$significance_agreement <- with(
  sensitivity_comparison,
  paired_significant == unpaired_significant
)

# Order the detailed table by the primary paired-model adjusted p-value.
sensitivity_comparison <- sensitivity_comparison[
  order(
    sensitivity_comparison$paired_interaction_padj,
    sensitivity_comparison$unpaired_interaction_padj,
    na.last = TRUE
  ),
]

write.csv(
  sensitivity_comparison,
  paste0(
    "DESeq2_results/DGE/",
    "AR_blockade_interaction_paired_vs_unpaired_all_genes.csv"
  ),
  row.names = FALSE
)

# Quantify agreement in effect sizes, directions, significant genes, and
# leading genes. Leading genes are the 20 smallest adjusted p-values in each
# model, regardless of whether they cross the 0.05 threshold.
finite_effects <- with(
  sensitivity_comparison,
  is.finite(paired_interaction_log2FC) &
    is.finite(unpaired_interaction_log2FC)
)

effect_spearman <- cor(
  sensitivity_comparison$paired_interaction_log2FC[finite_effects],
  sensitivity_comparison$unpaired_interaction_log2FC[finite_effects],
  method = "spearman"
)

overall_direction_agreement <- mean(
  sensitivity_comparison$direction_agrees[finite_effects],
  na.rm = TRUE
)

paired_sig_ids <- sensitivity_comparison$gene_id[
  sensitivity_comparison$paired_significant
]
unpaired_sig_ids <- sensitivity_comparison$gene_id[
  sensitivity_comparison$unpaired_significant
]

paired_sig_direction_agreement <- if (length(paired_sig_ids) > 0) {
  mean(
    sensitivity_comparison$direction_agrees[
      sensitivity_comparison$paired_significant
    ],
    na.rm = TRUE
  )
} else {
  NA_real_
}

top_n <- min(20, nrow(sensitivity_comparison))

paired_top_ids <- head(
  sensitivity_comparison$gene_id[
    order(sensitivity_comparison$paired_interaction_padj, na.last = TRUE)
  ],
  top_n
)

unpaired_top_ids <- head(
  sensitivity_comparison$gene_id[
    order(sensitivity_comparison$unpaired_interaction_padj, na.last = TRUE)
  ],
  top_n
)

shared_significant_ids <- intersect(paired_sig_ids, unpaired_sig_ids)
shared_top_ids <- intersect(paired_top_ids, unpaired_top_ids)

sensitivity_summary <- data.frame(
  metric = c(
    "Genes compared",
    "Spearman correlation of shrunken interaction log2FC",
    "Overall effect-direction agreement",
    "Significant genes in paired model (padj < 0.05)",
    "Significant genes in unpaired model (padj < 0.05)",
    "Significant genes shared by both models",
    "Direction agreement among paired-significant genes",
    paste0("Shared genes among each model's top ", top_n)
  ),
  value = c(
    sum(finite_effects),
    effect_spearman,
    overall_direction_agreement,
    length(paired_sig_ids),
    length(unpaired_sig_ids),
    length(shared_significant_ids),
    paired_sig_direction_agreement,
    length(shared_top_ids)
  ),
  stringsAsFactors = FALSE
)

write.csv(
  sensitivity_summary,
  paste0(
    "DESeq2_results/DGE/",
    "AR_blockade_interaction_paired_vs_unpaired_summary.csv"
  ),
  row.names = FALSE
)

message(
  "Paired vs unpaired interaction sensitivity analysis: Spearman rho = ",
  round(effect_spearman, 3),
  "; direction agreement = ",
  round(100 * overall_direction_agreement, 1),
  "%; significant genes paired/unpaired/shared = ",
  length(paired_sig_ids),
  "/",
  length(unpaired_sig_ids),
  "/",
  length(shared_significant_ids),
  "; shared top-",
  top_n,
  " genes = ",
  length(shared_top_ids),
  "."
)

print(sensitivity_summary)

# Effect-size agreement plot. Genes significant in the primary paired model
# are highlighted; the diagonal represents identical interaction estimates.
sensitivity_scatter <- ggplot(
  sensitivity_comparison,
  aes(
    x = paired_interaction_log2FC,
    y = unpaired_interaction_log2FC,
    color = paired_significant
  )
) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_hline(yintercept = 0, color = "grey80") +
  geom_vline(xintercept = 0, color = "grey80") +
  geom_point(alpha = 0.65, size = 1.5) +
  scale_color_manual(
    values = c("FALSE" = "grey70", "TRUE" = "#D7191C"),
    labels = c("FALSE" = "No", "TRUE" = "Yes")
  ) +
  labs(
    title = "Older tadpoles: paired vs unpaired interaction effects",
    subtitle = paste0(
      "Spearman rho = ",
      round(effect_spearman, 2),
      "; direction agreement = ",
      round(100 * overall_direction_agreement, 1),
      "%"
    ),
    x = "Paired interaction shrunken log2FC",
    y = "Unpaired interaction shrunken log2FC",
    color = "Paired padj < 0.05"
  ) +
  coord_equal() +
  theme_classic(base_size = 12) +
  theme(legend.position = "top")

ggsave(
  "DESeq2_results/QC/AR_blockade_interaction_paired_vs_unpaired.pdf",
  sensitivity_scatter,
  width = 8,
  height = 7
)

ggsave(
  "DESeq2_results/QC/AR_blockade_interaction_paired_vs_unpaired.png",
  sensitivity_scatter,
  width = 8,
  height = 7,
  dpi = 300
)

# ----------------------------
# 12. Identify paired-significant interaction genes
# ----------------------------

paired_significant_genes <- subset(
  interaction_out,
  !is.na(padj) & padj < 0.05
)

paired_significant_genes <- paired_significant_genes[, c(
  "gene_id",
  "gene_name",
  "product_description",
  "head_flutamide_log2FC",
  "head_flutamide_padj",
  "body_flutamide_log2FC",
  "body_flutamide_padj",
  "shrunken_interaction_log2FC",
  "padj",
  "interaction_result"
)]

paired_significant_genes <- paired_significant_genes[
  order(paired_significant_genes$padj),
]

print(paired_significant_genes, row.names = FALSE)

write.csv(
  paired_significant_genes,
  paste0(
    "DESeq2_results/DGE/",
    "AR_blockade_paired_significant_interaction_genes.csv"
  ),
  row.names = FALSE
)

# Genes significant in both paired and unpaired interaction models.
robust_interaction_genes <- subset(
  sensitivity_comparison,
  paired_significant & unpaired_significant
)

robust_interaction_genes <- robust_interaction_genes[, c(
  "gene_id",
  "gene_name",
  "product_description",
  "paired_interaction_log2FC",
  "paired_interaction_padj",
  "unpaired_interaction_log2FC",
  "unpaired_interaction_padj",
  "direction_agrees"
)]

write.csv(
  robust_interaction_genes,
  paste0(
    "DESeq2_results/DGE/",
    "AR_blockade_interaction_significant_in_both_models.csv"
  ),
  row.names = FALSE
)

# Genes significant only in the primary paired model.
paired_only_interaction_genes <- subset(
  sensitivity_comparison,
  paired_significant & !unpaired_significant
)

paired_only_interaction_genes <- paired_only_interaction_genes[, c(
  "gene_id",
  "gene_name",
  "product_description",
  "paired_interaction_log2FC",
  "paired_interaction_padj",
  "unpaired_interaction_log2FC",
  "unpaired_interaction_padj",
  "direction_agrees"
)]

write.csv(
  paired_only_interaction_genes,
  paste0(
    "DESeq2_results/DGE/",
    "AR_blockade_interaction_significant_in_paired_only.csv"
  ),
  row.names = FALSE
)

message(
  nrow(paired_significant_genes),
  " paired-significant interaction genes: ",
  nrow(robust_interaction_genes),
  " are also significant in the unpaired model and ",
  nrow(paired_only_interaction_genes),
  " are significant only in the paired model."
)

# Save the fitted object so analysis need not be rerun.
saveRDS(dds, "DESeq2_results/dds_fitted.rds")
saveRDS(vsd, "DESeq2_results/vsd.rds")
saveRDS(dds_old, "DESeq2_results/dds_old_interaction_fitted.rds")
saveRDS(
  dds_old_unpaired,
  "DESeq2_results/dds_old_unpaired_interaction_fitted.rds"
)

sessionInfo()
```

## Interpreting the older-only AR-blockade interaction

The primary paired analysis identified `r nrow(paired_significant_genes)` genes whose flutamide response differed significantly between head and body at an adjusted p-value threshold of 0.05. Of these, `r nrow(robust_interaction_genes)` were also significant in the unpaired sensitivity model, while `r nrow(paired_only_interaction_genes)` crossed the significance threshold only in the paired model.

The interaction is defined as:

$$
(\text{Flutamide head}-\text{Control head})-
(\text{Flutamide body}-\text{Control body}).
$$

A **positive interaction** means that the flutamide-associated change is more positive in head than in body. This could represent stronger induction in head, stronger repression in body, or both. A **negative interaction** means that the flutamide-associated change is more positive in body than in head; this could represent stronger induction in body, stronger repression in head, or both.

The interaction sign alone should therefore not be described as gene upregulation or downregulation. Interpret each gene using `head_flutamide_log2FC` and `body_flutamide_log2FC` alongside `shrunken_interaction_log2FC`:

- Head and body effects with the same sign but different magnitudes indicate a quantitatively tissue-biased response.
- Effects with opposite signs indicate a qualitatively tissue-dependent response and are especially strong candidates for follow-up.
- Genes significant in both paired and unpaired models have the strongest statistical robustness to the modeling choice.
- Genes significant only in the paired model may still be biologically meaningful, particularly when the unpaired effect has the same direction, but should be interpreted more cautiously.
- Statistical interaction does not establish direct androgen-receptor regulation. Changes may be indirect or may reflect altered developmental timing or tissue composition.
